---
title: "Regression models trail for Ridership and Gas Price"
author: "Rita Li"
date: '2022-10-31'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r,message=FALSE,echo=FALSE,warning=FALSE}
library(dbplyr)
library(tidyverse)
library(lubridate)
library(readr)
# for geeM studies
library(geeM)
library(tidyr)

library(dplyr)
library(broom)
library(ggplot2)
library(stringr)
library(splines)
library(astsa)


ridership<- read_csv("ridershipdata/ridership-route-day-Jan2014-Oct2017.csv")

gas <- read_csv("ridershipdata/Weekly_Minnesota_Regular_Conventional_Retail_Gasoline_Prices.csv", skip = 4)

nbhd <- read_csv("ridershipdata/stops_and_neighborhood.xls - TransitStops_Int_ExportTable.csv")
```

# prepare gas dataset
```{r, warning=FALSE}
gas_fixed <- gas %>% 
  mutate(date = as.Date(`Week of`, "%m/%d/%Y")) %>% 
  mutate(gas_price = `Weekly Minnesota Regular Conventional Retail Gasoline Prices Dollars per Gallon`) %>% 
  select(date, gas_price) %>% 
  mutate(month = month(date),year = year(date)) %>% 
  group_by(year,month) %>% 
  summarise(gas_price = mean(gas_price,na.rm = TRUE)) %>% 
  filter(year > 2013 & year < 2018) %>% 
  ungroup()

```

# prepare neighborhood dataset

```{r}
nbhd %>% 
  mutate(neighborhood = planning_1) %>% 
  distinct(route, routedscrp, neighborhood, RouteClass)
```

# Join Datasets
```{r, fig.width=7,fig.height=3,message=FALSE}
rider_gas_complete <- ridership %>% 
  mutate(date = as.Date(dtDate)) %>%
  mutate(month = month(date),year = year(date)) %>% 
  group_by(Route,year,month) %>% 
  summarise(riders = sum(Total_Riders),RouteType) %>% 
  unique() %>% 
  right_join(gas_fixed, by = c("month","year")) %>% 
  ungroup()

# join data
routeclass <- nbhd %>% 
  distinct(route, RouteClass)

coreloc <- rider_gas_complete %>% 
  right_join(routeclass, by = c("Route" = "route")) %>% 
  group_by(RouteClass, year, month) %>% 
  summarise(total_riders = mean(riders)) %>% 
  filter(RouteClass == "CoreLoc") %>% 
  mutate(date = year + (month-1)/12)
  
  
## I don't know
rider_gas_complete %>% 
  group_by(Route) %>% 
  mutate(gas_last_month = lag(gas_price, 1),
         change_from_last_month = gas_price-gas_last_month,
         per_change = change_from_last_month/gas_last_month,
         change = ifelse(per_change > 0, TRUE, FALSE)) %>% 
  mutate(ride_last_month = lag(riders, 1),
         ride_change = riders-ride_last_month,
         ride_per_change = ride_change/ride_last_month,
         rchange = ifelse(per_change > 0, TRUE, FALSE)) %>%
  filter(Route == 63) %>% 
  ggplot(aes(x = date, y = ride_per_change))+
  geom_line()+
  geom_line(aes(y = gas_price/5-1),color = "blue")
```

# failed models
```{r}
ride63 <- rider_gas_complete %>%
  drop_na(riders, date) %>% 
  mutate(date_num = year + (month-1)/12) %>% 
  count(RouteType)
  ungroup()

lm.season <- lm(riders ~ factor(month), data = ride63)

ride63%>% 
  filter(Route == 84|Route == 63) %>% 
  ggplot(aes(x = date, y = riders, color = factor(Route))) + 
  geom_line()+
  geom_line(aes(y = gas_price*100),color = "black")


lm(riders ~ gas_price+factor(Route), data = ride63) %>% 
  summary()

ride63 %>% 
  geem(riders ~ gas_price, data = ., id = Route, corstr = "ar1") %>% 
  summary()

gas <- ride63 %>% 
  filter(Route == 63)

r <- ride63 %>% 
  group_by(Route) %>% 
  mutate(schedules = n()) %>%
  ungroup() %>% 
  filter(schedules == max(schedules))
```
#CoreLoc

## timeseries for ridership
```{r}
coreloc %>% 
  mutate(date_num = year + (month-1)/12) %>%
  ggplot(aes(x = date_num, y = total_riders))+
  geom_line()
```
```{r}
acf2(na.omit(coreloc$total_riders)) 
```

```{r}
AR1 <- sarima(coreloc$total_riders, p = 1, d = 0, q = 0)
ARMA11 <- sarima(coreloc$total_riders, p = 1, d = 0, q = 1)
```
Result from the AR(1) model is not very perfect, which indicates that the change of the ridership of CoreLoc is not only random noise, there is other variables that have effect the ridership.

# time series for gas price
```{r}
gas_fixed %>% 
  mutate(date_num = year + (month-1)/12) %>%
  ggplot(aes(x = date_num, y = gas_price))+
  geom_line()
```
```{r}
lm.season <- lm(gas_price ~ factor(month), data = gas_fixed)

gas_fixed %>% 
  mutate(gas_predict = predict(lm.season)) %>% 
  mutate(date_num = year + (month-1)/12) %>%
  ggplot(aes(x = date_num, y = gas_price)) + 
  geom_line()+
  geom_line(aes(y = gas_predict),color = "red")

gas_fixed %>% 
  mutate(gas_predict = predict(lm.season)) %>% 
  mutate(date_num = year + (month-1)/12)%>%
  mutate(fixed_gas_price = gas_price-gas_predict) %>% 
  ggplot(aes(x = date_num, y = fixed_gas_price)) + 
  geom_line()
```
gas price that will used in the final model is fixed gas price which is calculated by "gas_price - gas_predidct".



### single vairable regression of each bus routes, riders and gas price, visualization of p-values
```{r}
routes <- r$Route %>% unique()
gasp <- gas$gas_price

betas <- c()
ses <- c()
tstats <- c()
pvals <- c()

# loop through all routes
a <- 0

for(i in routes){
  rides <- r %>% 
    filter(Route == i)
  ride <- rides$riders
  
  #fit the linear model
  mod <- lm(ride ~ gasp)
  # get coefficient information
  coefinfo <- tidy(mod)
  
  a <- a+1
  # record estimate, SE, test stat, and p-value
  betas[a] <- coefinfo$estimate[2]
  ses[a] <- coefinfo$std.error[2]
  tstats[a] <- coefinfo$statistic[2]
  pvals[a] <- coefinfo$p.value[2]
}

mod.results <- routes %>%
  as.data.frame() %>% 
  mutate(Estimate = betas,
         Std.Error = ses,
         Test.Statistic = tstats,
         P.Value = pvals)

mod.results %>% 
  mutate(sig = ifelse(P.Value <= 0.01,TRUE,FALSE)) %>% 
  mutate(positive = ifelse(Estimate > 0, TRUE, FALSE)) %>% 
  ggplot(aes(x = ., y = log(P.Value),shape = sig))+
  geom_point(aes(color = positive))
```

### zscore of ridership and gas price linear regression and prediction visualization
```{r}
justed_data <- rider_gas_complete %>% 
  group_by(Route) %>% 
  mutate(avg_riders = mean(riders),
         justed_riders = (riders-avg_riders)/sd(riders)) %>% 
  mutate(schedules = n()) %>%
  ungroup() %>% 
  filter(schedules == 44)

mod.justed <- justed_data %>% 
  lm(justed_riders ~ gas_price, data = .)

mod.justed %>% summary()

justed_data %>% 
  mutate(predict = predict(mod.justed),
         predict_riders = predict * sd(riders) + avg_riders) %>% 
  filter(Route == 63) %>% 
  ggplot(aes(x = date, y = riders))+
  geom_line()+
  geom_line(aes(y = predict_riders), color = "blue")
```

### zscore and gas price regression for each bus routes and visualization of p-values.
```{r}
routes <- justed_data$Route %>% unique()
gasp <- gas$gas_price

#gasp[1] <- 0

betas <- c()
ses <- c()
tstats <- c()
pvals <- c()

# loop through all routes
a <- 0

for(i in routes){
  rides <- justed_data %>% 
    filter(Route == i)
  ride <- rides$justed_riders
  
  #fit the linear model
  mod <- lm(ride ~ gasp)
  # get coefficient information
  coefinfo <- tidy(mod)
  
  a <- a+1
  # record estimate, SE, test stat, and p-value
  betas[a] <- coefinfo$estimate[2]
  ses[a] <- coefinfo$std.error[2]
  tstats[a] <- coefinfo$statistic[2]
  pvals[a] <- coefinfo$p.value[2]
}

mod.results <- routes %>%
  as.data.frame() %>% 
  mutate(Estimate = betas,
         Std.Error = ses,
         Test.Statistic = tstats,
         P.Value = pvals)

mod.results %>% 
  mutate(sig = ifelse(P.Value <= 0.05,TRUE,FALSE)) %>% 
  mutate(positive = ifelse(Estimate > 0, TRUE, FALSE)) %>% 
  ggplot(aes(x = ., y = log(P.Value),shape = sig))+
  geom_point(aes(color = positive))

```