---
title: "Regression models trail for Ridership and Gas Price"
author: "Rita Li, Jenny Li"
date: '2022-10-31'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```
# Research Question and Motivation
We have two research questions, both mainly focusing on the relationship between gas price and ridership. The two research questions are: What is the relationship between gas price and ridership of bus? And how is this relationship different within different types of bus routes? Our motivation for those research questions is due to constantly increasing gas prices in recent years. We are curious how companies,like metro transit, or policy makers would change their strategy to deal with such a situation.

# Data Set Up

## Load Library 
```{r,message=FALSE,echo=FALSE,warning=FALSE}
library(dbplyr)
library(tidyverse)
library(lubridate)
library(readr)
# for geeM studies
library(geeM)
library(tidyr)

library(dplyr)
library(broom)
library(ggplot2)
library(stringr)
library(splines)
library(astsa)

```

## Datasets Read In and Their Description
```{r data read in}
# Ridership data. Ridership our dependent variable in the research question. This dataset on ridership woudl provide us information on the ridership which enables us to build model from it.
ridership<- read_csv("ridershipdata/ridership-route-day-Jan2014-Oct2017.csv")

# Gas price data. Gas price data is our main independent variable in the research question. This dataset on gas price provide us the trend of gas price which could be compared with the ridership.
gas <- read_csv("ridershipdata/Weekly_Minnesota_Regular_Conventional_Retail_Gasoline_Prices.csv", skip = 4)

#Routes and neighborhood data. This dataset provide information of transit system. Including the route types route ID and other useful information for comparing their responding ridership.
nbhd <- read_csv("ridershipdata/stops_and_neighborhood.xls - TransitStops_Int_ExportTable.csv")
```

## Prepare Gas Dataset
```{r, warning=FALSE}
# Calculate the mean for gas price in each month each year
gas_fixed <- gas %>% 
  mutate(date = as.Date(`Week of`, "%m/%d/%Y")) %>% 
  mutate(gas_price = `Weekly Minnesota Regular Conventional Retail Gasoline Prices Dollars per Gallon`) %>% 
  select(date, gas_price) %>% 
  mutate(month = month(date),year = year(date)) %>% 
  group_by(year,month) %>% 
  summarise(gas_price = mean(gas_price,na.rm = TRUE)) %>% 
  #filter(year > 2013 & year < 2018) %>% 
  ungroup() %>% 
  mutate(date = year + (month-1)/12)
```

## Pepare Route Class Dataset

```{r}
#Clean the routes data
routeclass <- nbhd %>% 
  distinct(route, RouteClass)
```

## Join Datasets
```{r, fig.width=7,fig.height=3,message=FALSE}
# Combine the ridership and the gas price
rider_gas_complete <- ridership %>% 
  mutate(date = as.Date(dtDate)) %>%
  mutate(month = month(date),year = year(date)) %>% 
  group_by(Route,year,month) %>% 
  summarise(riders = sum(Total_Riders),RouteType) %>% 
  unique() %>% 
  right_join(gas_fixed, by = c("month","year")) %>% 
  ungroup()


#Mean ridership data for CoreLoc(one of the bus route types)
coreloc <- rider_gas_complete %>% 
  right_join(routeclass, by = c("Route" = "route")) %>% 
  group_by(RouteClass, year, month) %>% 
  summarise(total_riders = mean(riders)) %>% 
  filter(RouteClass == "CoreLoc") %>% 
  mutate(date = year + (month-1)/12)
```

# Models

## CoreLoc

### Visualization of CoreLoc Ridership

```{r, fig.width=7,fig.height=3,message=FALSE}
coef <- 10000
CoreLoc %>% 
  ggplot(aes(x = date))+
  geom_line(aes(y = gas_price),color = "#9F2C2C")+
  geom_line(aes(y = total_riders/coef-10),color = "#3F4345")+ 
  scale_y_continuous(
    name = "Gas Price($)",
    sec.axis = sec_axis(trans = ~.*coef-10, name="Total Riders(thousands)"))+
  labs(title = "Change of Gas Price and Total Rides of all CoreLoc Routes over time",
       subtitle = "From Jan 2014 to Oct 2017")+
  geom_text(aes(x = 2017.5, y = 2.4), label = "Gas Price",color = "#9F2C2C",size = 3)+
  geom_text(aes(x = 2017.5, y = 1.7), label = "Total Riders",color = "#3F4345",size = 3)+
  theme_minimal()+
  theme(axis.title.y.left = element_text(color = "#9F2C2C",size = 12),
        axis.title.y.right = element_text(color = "#3F4345",size = 12),
        axis.text.y.left = element_text(color = "#9F2C2C",size = 8),
        axis.text.y.right = element_text(color = "#3F4345",size = 8),
        axis.title.x.bottom = element_blank(),
        plot.title = element_text(color = "#006bb3",size = 14),
        plot.subtitle = element_text(color = "#006bb3"))
```

### Fit factor(month) + gas_price Model for CoreLoc
```{r}
CoreLoc <- rider_gas_complete %>% 
  right_join(routeclass, by = c("Route" = "route")) %>% 
  group_by(RouteClass, year, month) %>% 
  summarise(total_riders = mean(riders),gas_price) %>% 
  filter(RouteClass == "CoreLoc") %>% 
  mutate(date = year + (month-1)/12) %>% 
  drop_na() %>%
  unique() %>% 
  ungroup()

lm.ride.season <- lm(total_riders ~ factor(month) + gas_price, data = CoreLoc)

CoreLoc$ride_predict <- predict(lm.ride.season)

CoreLoc$remove_season <- CoreLoc$total_riders-CoreLoc$ride_predict

lm.ride.season %>% summary()
```
```{r,fig.height=2,fig.width=4}
CoreLoc %>% 
  ggplot(aes(x = date, y = total_riders)) + 
  geom_line()+
  geom_line(aes(y = ride_predict),color = "blue",alpha = 0.3)+
  theme_minimal()+
  labs(title = "Time series of ridership of CoreLoc",
       x = "Date(numeric)",
       y = "Total Riders")
```

### Timeseries for CoreLoc Ridership

```{r}
# look at residuals
CoreLoc %>% 
  ggplot(aes(x = date, y = remove_season))+
  geom_line()+
  theme_minimal()
```

```{r}
acf2(na.omit(CoreLoc$remove_season)) 
```

If we take a look at the ACF(auto correlation function) and PACF(partial auto-correlation function) as shown below, we can see that the ACF gradually decays to 0, and the PACF seems to drop to 0 after lag 3. The patterns we observe here match the theoretical pattern of an AR model as the ACF gradually drops to 0 and the PACF seems to drop to 0 after a certain lag. Since the PACF seems to drop to 0 around after lag 2, we decided to fit an AR(3) model for the residuals.

```{r}
AR2 <- sarima(CoreLoc$remove_season, p = 3, d = 0, q = 0)
```
We think our model does a decent job at estimating our data as the ACF of residuals are fairly small, and the p-values for Ljung-Box test are fairly large, which implies our residuals are independent. 


## CommExpress

### Visualization
```{r, fig.width=7,fig.height=3,message=FALSE}
coef <- 4000
CommExpress %>% 
  # mutate(gas_price_normalized = (gas_price - mean(gas_price)/sd(gas_price)),
  #        riders_normalized = (total_riders - mean(total_riders))/sd(total_riders)) %>% 
  ggplot(aes(x = date))+
  geom_line(aes(y = gas_price),color = "#9F2C2C")+
  geom_line(aes(y = total_riders/coef-1),color = "#3F4345")+ 
  scale_y_continuous(
    name = "Gas Price($)",
    sec.axis = sec_axis(trans = ~.*coef-1, name="Total Riders"))+
  labs(title = "Change of Gas Price and Total Rides of all CommExpress Routes over time",
       subtitle = "From Jan 2014 to Oct 2017")+
  geom_text(aes(x = 2017.5, y = 2.1), label = "Gas Price",color = "#9F2C2C",size = 3)+
  geom_text(aes(x = 2017.5, y = 2.85), label = "Total Riders",color = "#3F4345",size = 3)+
  theme_minimal()+
  theme(axis.title.y.left = element_text(color = "#9F2C2C",size = 12),
        axis.title.y.right = element_text(color = "#3F4345",size = 12),
        axis.text.y.left = element_text(color = "#9F2C2C",size = 8),
        axis.text.y.right = element_text(color = "#3F4345",size = 8),
        axis.title.x.bottom = element_blank(),
        plot.title = element_text(color = "#006bb3",size = 14),
        plot.subtitle = element_text(color = "#006bb3"))
```

### Fit factor(month) + gas_price Model for CommExpress
```{r}
CommExpress <- rider_gas_complete %>% 
  right_join(routeclass, by = c("Route" = "route")) %>% 
  group_by(RouteClass, year, month) %>% 
  summarise(total_riders = mean(riders),gas_price) %>% 
  filter(RouteClass == "CommExpress") %>% 
  mutate(date = year + (month-1)/12) %>% 
  drop_na() %>% 
  unique() %>% 
  ungroup()

lm.ride.season <- lm(total_riders ~ factor(month) + gas_price, data = CommExpress)

CommExpress$ride_predict <- predict(lm.ride.season)

CommExpress$remove_season <- CommExpress$total_riders-CommExpress$ride_predict

lm.ride.season %>% summary()
```



## Other Models

### Single Vairable Regression of Each Bus Routes, Riders and Gas Price, Visualization of P-values
```{r}
routes <- r$Route %>% unique()
gasp <- gas$gas_price

betas <- c()
ses <- c()
tstats <- c()
pvals <- c()

# loop through all routes
a <- 0

for(i in routes){
  rides <- r %>% 
    filter(Route == i)
  ride <- rides$riders
  
  #fit the linear model
  mod <- lm(ride ~ gasp)
  # get coefficient information
  coefinfo <- tidy(mod)
  
  a <- a+1
  # record estimate, SE, test stat, and p-value
  betas[a] <- coefinfo$estimate[2]
  ses[a] <- coefinfo$std.error[2]
  tstats[a] <- coefinfo$statistic[2]
  pvals[a] <- coefinfo$p.value[2]
}

mod.results <- routes %>%
  as.data.frame() %>% 
  mutate(Estimate = betas,
         Std.Error = ses,
         Test.Statistic = tstats,
         P.Value = pvals)

mod.results %>% 
  mutate(sig = ifelse(P.Value <= 0.01,TRUE,FALSE)) %>% 
  mutate(positive = ifelse(Estimate > 0, TRUE, FALSE)) %>% 
  ggplot(aes(x = ., y = log(P.Value),shape = sig))+
  geom_point(aes(color = positive))
```

### Z-score of Ridership and Gas Price Linear Regression and Prediction Visualization
```{r}
justed_data <- rider_gas_complete %>% 
  group_by(Route) %>% 
  mutate(avg_riders = mean(riders),
         justed_riders = (riders-avg_riders)/sd(riders)) %>% 
  mutate(schedules = n()) %>%
  ungroup() %>% 
  filter(schedules == 44)

mod.justed <- justed_data %>% 
  lm(justed_riders ~ gas_price, data = .)

mod.justed %>% summary()

justed_data %>% 
  mutate(predict = predict(mod.justed),
         predict_riders = predict * sd(riders) + avg_riders) %>% 
  filter(Route == 63) %>% 
  ggplot(aes(x = date, y = riders))+
  geom_line()+
  geom_line(aes(y = predict_riders), color = "blue")
```

### Z-score and Gas Price Regression for Each Bus Routes and Visualization of P-values.
```{r}
routes <- justed_data$Route %>% unique()
gasp <- gas$gas_price

#gasp[1] <- 0

betas <- c()
ses <- c()
tstats <- c()
pvals <- c()

# loop through all routes
a <- 0

for(i in routes){
  rides <- justed_data %>% 
    filter(Route == i)
  ride <- rides$justed_riders
  
  #fit the linear model
  mod <- lm(ride ~ gasp)
  # get coefficient information
  coefinfo <- tidy(mod)
  
  a <- a+1
  # record estimate, SE, test stat, and p-value
  betas[a] <- coefinfo$estimate[2]
  ses[a] <- coefinfo$std.error[2]
  tstats[a] <- coefinfo$statistic[2]
  pvals[a] <- coefinfo$p.value[2]
}

mod.results <- routes %>%
  as.data.frame() %>% 
  mutate(Estimate = betas,
         Std.Error = ses,
         Test.Statistic = tstats,
         P.Value = pvals)

mod.results %>% 
  mutate(sig = ifelse(P.Value <= 0.05,TRUE,FALSE)) %>% 
  mutate(positive = ifelse(Estimate > 0, TRUE, FALSE)) %>% 
  ggplot(aes(x = ., y = log(P.Value),shape = sig))+
  geom_point(aes(color = positive))

```

## Failed Models
```{r, eval=FALSE}
ride63 <- rider_gas_complete %>%
  drop_na(riders, date) %>% 
  mutate(date_num = year + (month-1)/12) %>% 
  count(RouteType)
  ungroup()

lm.season <- lm(riders ~ factor(month), data = ride63)

ride63%>% 
  filter(Route == 84|Route == 63) %>% 
  ggplot(aes(x = date, y = riders, color = factor(Route))) + 
  geom_line()+
  geom_line(aes(y = gas_price*100),color = "black")


lm(riders ~ gas_price+factor(Route), data = ride63) %>% 
  summary()

ride63 %>% 
  geem(riders ~ gas_price, data = ., id = Route, corstr = "ar1") %>% 
  summary()

gas <- ride63 %>% 
  filter(Route == 63)

r <- ride63 %>% 
  group_by(Route) %>% 
  mutate(schedules = n()) %>%
  ungroup() %>% 
  filter(schedules == max(schedules))
```

# Visualization

## Shiny App

[shiny app](https://jennylipro.shinyapps.io/routesMap/)

```{r,eval=FALSE}
#library
library(shiny)
library(shinyWidgets)
library(rgdal)
library(leaflet)
library (raster)
library(dplyr)
library(sf)
library(tidyverse)
#load data
map <- readOGR("newmap.shp") %>%
  st_as_sf()

#Class name for checkbox
className <- map %>%
  distinct(RouteClass) %>%
  mutate(dum = "1") %>%
  pivot_wider(names_from = RouteClass, values_from = dum)

mychoice = names(className)


ui <- fluidPage(# Application title
  titlePanel("Twin Cities Metro Route Map"),
  
  sidebarLayout(
    sidebarPanel(
      h4("Please choose route classes to include :"),
      #Select ALL
      checkboxInput("all", "Select All/None"),
      #add checkbox group
      checkboxGroupInput("classSelect", "Route Classes", mychoice)
      
      # #checkbox list of route class
      # uiOutput("classes_select "),
    ),
    # Show a plot of the generated distribution
    mainPanel(#Dropdown list of routes
      uiOutput("routes_select"),
      
      
      #Map
      leafletOutput("routesLeaflet"),)
  ))

# Define server logic required to draw a histogram
server <- function(input, output, session) {
  #update checkbox input
  observe({
    updateCheckboxGroupInput(session,
                             "classSelect",
                             choices = mychoice,
                             #if select all is true, then show all the choices, vice versa
                             selected = if (input$all)
                               mychoice)
  })
  
  #creat routes dropdown list
  output$routes_select <- renderUI({
    #we want the routes name only appears once
    selectInput("pick_routes", "choose a route:", choices = c(unique(map$route)))
  })
  
  
  #create map
  output$routesLeaflet <- renderLeaflet({
    #make it only showed the route selected
    filtered = filter(map, map$route == input$pick_routes)
    #class filter
    filteredClass = filter(map, map$RouteClass %in%  c(input$classSelect))
    #set up color
    factpal <- colorFactor(palette = "Set1", filteredClass$RouteClass)
    #setup map
    routesLeaflet <-
      #map%>%
      leaflet() %>%
      addProviderTiles("CartoDB.Positron",
                       options = providerTileOptions(opacity = 0.8)) %>%
      addPolylines(
        data = filtered,
        color = "navy",
        weight = 3,
        smoothFactor = 0.5,
        opacity = 1.0,
        highlightOptions = highlightOptions(
          color = "cyan",
          weight = 2,
          bringToFront = TRUE
        ),
        popup =  paste("Route: ", map$route, "<br>")
      ) %>%
      addPolylines(
        data = filteredClass,
        color = ~ factpal(RouteClass),
        weight = 1,
        smoothFactor = 0.5,
        opacity = 0.5,
        highlightOptions = highlightOptions(
          color = "yellow",
          weight = 2,
          bringToFront = TRUE
        ),
        popup =  paste("Route: ", map$route, "<br>")
      )
    
    
    #call map
    routesLeaflet
  })
  
}

# Run the application
shinyApp(ui = ui, server = server)

```

# Member Contribution
