---
title: "dataset for single route"
author: "Rita Li"
date: '2022-12-02'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,message=FALSE,echo=FALSE,warning=FALSE}
library(dbplyr)
library(tidyverse)
library(lubridate)
library(readr)
library(tidyr)

library(dplyr)
library(broom)
library(ggplot2)
library(stringr)
library(splines)
library(astsa)

library(rjson)
library(geojsonsf)
library(sf)
library(rgdal)

library(zoo)
```

```{r}
## Original ridership dataset
ridership<- read_csv("ridershipdata/ridership-route-day-Jan2014-Oct2017.csv") %>% 
  mutate(date = as.Date(dtDate)) %>% 
  select(Route, date, Total_Riders, Schedule, SumOfTrips)

## ridership of 10 highest riders routes from 2008 to 2019, combined to previous dataset later
high_route_ridership <- read_csv("high_route_ridership.csv")

## weekly gas dataset
gas <- read_csv("ridershipdata/Weekly_Minnesota_Regular_Conventional_Retail_Gasoline_Prices.csv", skip = 4) %>% 
  mutate(date = as.Date(`Week of`, "%m/%d/%Y")) %>% 
  mutate(gas_price = `Weekly Minnesota Regular Conventional Retail Gasoline Prices Dollars per Gallon`) %>% 
  select(date, gas_price) %>% 
  mutate(month = month(date),year = year(date)) %>% 
  group_by(year,month) %>% 
  summarise(gas_price = mean(gas_price,na.rm = TRUE)) %>% 
  ungroup()

## filter the ridership data to all routes that are also in the shapefile(still working in 2022)
map <- geojson_sf("shapedata/TransitRoutes.json",expand_geometries = TRUE)

RouteClass <- map$RouteClass
RRoute <- map$route
dscp <- map$routedscrp

routeclass <- cbind(RouteClass, RRoute, dscp) %>% 
  as.data.frame() %>% 
  mutate(route = as.numeric(RRoute))

## 69 routes are in both ridership dataset and shapefile, 64 of them have at least two year data
```

```{r, fig.width=7,fig.height=3,message=FALSE}
# Combine the ridership and the gas price
rider_gas_complete <- ridership %>%
  full_join(high_route_ridership, by = c("Route" = "route","date")) %>% 
  inner_join(routeclass, by = c("Route" = "route")) %>%
  mutate(Total_Riders = ifelse(is.na(Total_Riders), rides, Total_Riders),
         month = month(date),year = year(date),
         weekday = ifelse(Schedule == "Weekday",1,0)) %>% 
  group_by(Route,year,month) %>% 
  mutate(wd =  sum(weekday),wd2 = n()) %>% 
  summarise(riders = sum(Total_Riders,na.rm = TRUE),
            date = min(date),
            wd,wd2) %>% 
  ungroup() %>% 
  unique() %>% 
  inner_join(routeclass, by = c("Route" = "route")) %>%
  mutate(NumOfWdays = ifelse(RouteClass == "CoreLoc", wd2, wd),
         avg_riders = riders/NumOfWdays) %>% 
  left_join(gas, by = c("month","year")) %>% 
  ungroup() %>% 
  group_by(Route) %>% 
  mutate(n = n()) %>%
  ungroup() %>% 
  filter(n >= 24)

rider_gas_complete %>% 
  filter(RouteClass == "CoreLoc") %>% 
  count(year,month,NumOfWdays)

# write_csv(rider_gas_complete, file = "ridershipdata/rider_gas_single_route.csv")
```


```{r}
rider_gas_complete <- read_csv("ridershipdata/rider_gas_single_route.csv")


# missing value for LRT(901 & 902) at Sept 2015 and Oct 2016, replaced with rolling mean
rider_gas_complete <- rider_gas_complete %>% 
  mutate(previous = lag(riders, n = 1L),
         nextt = lead(riders, n = 1L),
         riders = ifelse(riders == 0, (previous+nextt)/2,riders)) %>% 
  select(-previous, -nextt) %>% 
  mutate(previous = lag(avg_riders, n = 1L),
         nextt = lead(avg_riders, n = 1L),
         avg_riders = ifelse(avg_riders == 0, (previous+nextt)/2,avg_riders)) %>% 
  select(-previous, -nextt) %>% 
  filter(!(Route == 888 & date > "2015-06-30"))
  
  rider_gas_complete %>% 
    filter(RouteClass == "CoreLoc") %>% 
    count(Route)
```



```{r}
routes <- rider_gas_complete$Route %>% 
  unique()

de_season <- data.frame()

for(i in routes){
  rides <- rider_gas_complete %>% 
    filter(Route == i)
  # ride <- rides$riders
  
  #fit the linear model
  mod <- lm(riders ~ factor(month), na.action=na.exclude, data = rides)
  # get coefficient information
  coefinfo <- tidy(mod)
  
  pre_riders <- rides %>% 
    mutate(prediction = predict(mod),
           riders = ifelse(is.na(riders), prediction, riders),
           remove_seasonality = riders - prediction) %>% 
    select(Route, year, month, prediction, remove_seasonality)
  
  # binding dataset
  de_season <- rbind(de_season, pre_riders)
  
  # a <- a+1
  # # record estimate, SE, test stat, and p-value
  # betas[a] <- coefinfo$estimate[2]
  # ses[a] <- coefinfo$std.error[2]
  # tstats[a] <- coefinfo$statistic[2]
  # pvals[a] <- coefinfo$p.value[2]
  # cors[a]<- cor(rides$riders, rides$gas_price)
}

de_season %>% 
  # filter(remove_seasonality < -200000) %>%
  filter(Route<900) %>% 
  mutate(date = paste(year,month,"01",sep="/")) %>% 
  ggplot(aes(x = date, y = remove_seasonality, group = Route)) +
  geom_line(aes(color = Route))
```

```{r}
de_season1<- data.frame()

for(i in routes){
  rides <- rider_gas_complete %>% 
    filter(Route == i)
  # ride <- rides$riders
  
  #fit the linear model
  mod <- lm(avg_riders ~ factor(month), na.action=na.exclude, data = rides)
  # get coefficient information
  coefinfo <- tidy(mod)
  
  pre_riders <- rides %>% 
    mutate(prediction = predict(mod),
           riders = ifelse(is.na(riders), prediction, riders),
           remove_seasonality = avg_riders - prediction) %>% 
    select(Route, year, month, prediction, remove_seasonality)
  
  # binding dataset
  de_season1 <- rbind(de_season1, pre_riders)
  
  # a <- a+1
  # # record estimate, SE, test stat, and p-value
  # betas[a] <- coefinfo$estimate[2]
  # ses[a] <- coefinfo$std.error[2]
  # tstats[a] <- coefinfo$statistic[2]
  # pvals[a] <- coefinfo$p.value[2]
  # cors[a]<- cor(rides$riders, rides$gas_price)
}

de_season1 %>% 
  # filter(remove_seasonality < -200000) %>%
  filter(Route<900) %>% 
  mutate(date = paste(year,month,"01",sep="/")) %>% 
  ggplot(aes(x = date, y = remove_seasonality, group = Route)) +
  geom_line(aes(color = Route))
```

```{r}
# for graph: norm_de_sea: normalized remove-season daily riders, use `remove_seasonality` for label/pop-up;
# norm_avg_riders: normalized original daily riders, use `avg_riders` for label/pop-up


rider_gas_complete <- rider_gas_complete %>% 
  left_join(de_season1, by = c("Route", "year","month")) %>% 
  #normalize de_season
  group_by(Route) %>% 
  mutate(meann = mean(remove_seasonality),
         norm_de_sea = (remove_seasonality-meann)/sd(remove_seasonality)/2+1) %>% 
  mutate(meann = mean(riders),
         norm_avg_riders = (riders-meann)/sd(riders)/2.5+1.5) %>%
  select(-meann,-wd,-wd2, -prediction,-NumOfWdays,-RRoute,-n) %>% 
  ungroup()

final_single_route <- rider_gas_complete %>% 
  mutate(`Date` = paste0(month.abb[month]," ", year,"\nGas Price: ", gas_price),
         `Daily Riders`= round(avg_riders, 0),
         `Daily Riders(removed seasonality)` = round(remove_seasonality,0)) %>% 
  mutate(`Daily Riders and Gas Price Line Plot` = norm_avg_riders,
         `Daily Riders(removed seasonality) and Gas Price Line Plot` = norm_de_sea)

write_csv(final_single_route, file = "multiple_plot_with_leaflet_demo/single_route.csv")
```


```{r}
rider_gas_complete %>% 
  left_join(de_season, by = c("Route", "year","month")) %>% 
  
  #normalize de_season
  group_by(Route) %>% 
  mutate(meann = mean(remove_seasonality),
         norm_de_sea = (remove_seasonality-meann)/sd(remove_seasonality)/2+1) %>% 
  # filter(Route == 114) %>% 
  filter(year > 2014 & year < 2018) %>% 
  filter(RouteClass == "CoreLoc") %>% 
  ggplot(aes(x = date))+
  geom_line(aes(y = gas_price),color = "#9F2C2C")+
  # geom_line(aes(y = remove_seasonality/coef),color = "#3F4345")+ 
  geom_line(aes(y = norm_de_sea, group = Route, color = Route))+
  scale_y_continuous(
    name = "Gas Price($)",
    sec.axis = sec_axis(trans = ~., name="Total Riders(thousands)"))+
  labs(title = "Change of Gas Price and Total Rides of Route 63 over time",
       subtitle = "From Jan 2014 to Oct 2017")+
  geom_text(aes(x = date("2017-06-01"), y = 2.4), label = "Gas Price",color = "#9F2C2C",size = 3)+
  geom_text(aes(x = date("2017-06-01"), y = 1.7), label = "Total Riders",color = "#3F4345",size = 3)+
  theme_minimal()+
  theme(axis.title.y.left = element_text(color = "#9F2C2C",size = 12),
        axis.title.y.right = element_text(color = "#3F4345",size = 12),
        axis.text.y.left = element_text(color = "#9F2C2C",size = 8),
        axis.text.y.right = element_text(color = "#3F4345",size = 8),
        axis.title.x.bottom = element_blank(),
        plot.title = element_text(color = "#006bb3",size = 14),
        plot.subtitle = element_text(color = "#006bb3"))

rider_gas_complete %>% 
  left_join(de_season1, by = c("Route", "year","month")) %>% 
  #normalize de_season
  group_by(Route) %>% 
  mutate(meann = mean(remove_seasonality),
         norm_de_sea = (remove_seasonality-meann)/sd(remove_seasonality)/2+1) %>% 
  # filter(Route == 114) %>% 
  filter(year > 2014 & year < 2018) %>% 
  filter(RouteClass == "CoreLoc") %>% 
  ggplot(aes(x = date))+
  geom_line(aes(y = gas_price),color = "#9F2C2C")+
  # geom_line(aes(y = remove_seasonality/coef),color = "#3F4345")+ 
  geom_line(aes(y = norm_de_sea, group = Route, color = Route))+
  scale_y_continuous(
    name = "Gas Price($)",
    sec.axis = sec_axis(trans = ~., name="Total Riders(thousands)"))+
  labs(title = "Change of Gas Price and Total Rides of Route 63 over time",
       subtitle = "From Jan 2014 to Oct 2017")+
  geom_text(aes(x = date("2017-06-01"), y = 2.4), label = "Gas Price",color = "#9F2C2C",size = 3)+
  geom_text(aes(x = date("2017-06-01"), y = 1.7), label = "Total Riders",color = "#3F4345",size = 3)+
  theme_minimal()+
  theme(axis.title.y.left = element_text(color = "#9F2C2C",size = 12),
        axis.title.y.right = element_text(color = "#3F4345",size = 12),
        axis.text.y.left = element_text(color = "#9F2C2C",size = 8),
        axis.text.y.right = element_text(color = "#3F4345",size = 8),
        axis.title.x.bottom = element_blank(),
        plot.title = element_text(color = "#006bb3",size = 14),
        plot.subtitle = element_text(color = "#006bb3"))
```


```{r}
rider_gas_complete <- read_csv("finaldatasets/single_route.csv")

GRAPH1 <- rider_gas_complete %>% 
  ungroup() %>% 
  mutate(`Date` = paste0(month.abb[month]," ", year,"\nGas Price: ", gas_price),
         `Daily Riders`= round(avg_riders, 0)) %>% 
  filter(Route == 63) %>% 
  ggplot(aes(x = date))+
  geom_line(aes(y = gas_price,label = `Date`),color = "#9F2C2C")+
  geom_line(aes(y = norm_de_sea, label = `Daily Riders`))+
  scale_y_continuous(
    name = "Gas Price($)",
    sec.axis = sec_axis(trans = ~
                          \., name="Total Riders(thousands)"))+
  labs(title = "Change of Gas Price and Total Rides of Route 63 over time",
       subtitle = "From Jan 2014 to Oct 2017")+
  geom_text(aes(x = date("2017-06-01"), y = 2.4), label = "Gas Price",color = "#9F2C2C",size = 3)+
  geom_text(aes(x = date("2017-06-01"), y = 1.7), label = "Total Riders",color = "#3F4345",size = 3)+
  theme_minimal()+
  theme(axis.title.y.left = element_text(color = "#9F2C2C",size = 12),
        axis.title.y.right = element_text(color = "#3F4345",size = 12),
        axis.text.y.left = element_text(color = "#9F2C2C",size = 8),
        axis.text.y.right = element_text(color = "#3F4345",size = 8),
        axis.title.x.bottom = element_blank(),
        plot.title = element_text(color = "#006bb3",size = 14),
        plot.subtitle = element_text(color = "#006bb3"))

# library(plotly)

ggplotly(GRAPH1, tooltip = c("label")) %>% 
   layout(hovermode = "x unified")
```




rider_gas_complete %>% 
  # left_join(de_season, by = c("Route", "year","month")) %>% 
  #normalize de_season
  group_by(Route) %>% 
  mutate(meann = mean(riders),
         norm_de_sea = (riders-meann)/sd(riders)/2+1) %>% 
  # filter(Route == 114) %>% 
  filter(year > 2014 & year < 2018) %>% 
  # filter(RouteClass == "CommExpress") %>% 
  ggplot(aes(x = date))+
  geom_line(aes(y = gas_price),color = "#9F2C2C")+
  # geom_line(aes(y = remove_seasonality/coef),color = "#3F4345")+ 
  geom_line(aes(y = norm_de_sea, group = Route, color = Route))+
  scale_y_continuous(
    name = "Gas Price($)",
    sec.axis = sec_axis(trans = ~.*coef, name="Total Riders(thousands)"))+
  labs(title = "Change of Gas Price and Total Rides of Route 63 over time",
       subtitle = "From Jan 2014 to Oct 2017")+
  geom_text(aes(x = date("2017-06-01"), y = 2.4), label = "Gas Price",color = "#9F2C2C",size = 3)+
  geom_text(aes(x = date("2017-06-01"), y = 1.7), label = "Total Riders",color = "#3F4345",size = 3)+
  theme_minimal()+
  theme(axis.title.y.left = element_text(color = "#9F2C2C",size = 12),
        axis.title.y.right = element_text(color = "#3F4345",size = 12),
        axis.text.y.left = element_text(color = "#9F2C2C",size = 8),
        axis.text.y.right = element_text(color = "#3F4345",size = 8),
        axis.title.x.bottom = element_blank(),
        plot.title = element_text(color = "#006bb3",size = 14),
        plot.subtitle = element_text(color = "#006bb3"))+
  facet_wrap(~RouteClass)

rider_gas_complete %>% 
  # left_join(de_season, by = c("Route", "year","month")) %>% 
  #normalize de_season
  group_by(Route) %>% 
  mutate(meann = mean(avg_riders),
         norm_de_sea = (avg_riders-meann)/sd(avg_riders)/2+1) %>% 
  # filter(Route == 114) %>% 
  filter(year > 2014 & year < 2018) %>% 
  # filter(RouteClass == "CommExpress") %>% 
  ggplot(aes(x = date))+
  geom_line(aes(y = gas_price),color = "#9F2C2C")+
  # geom_line(aes(y = remove_seasonality/coef),color = "#3F4345")+ 
  geom_line(aes(y = norm_de_sea, group = Route, color = Route))+
  scale_y_continuous(
    name = "Gas Price($)",
    sec.axis = sec_axis(trans = ~., name="Total Riders(thousands)"))+
  labs(title = "Change of Gas Price and Total Rides of Route 63 over time",
       subtitle = "From Jan 2014 to Oct 2017")+
  geom_text(aes(x = date("2017-06-01"), y = 2.4), label = "Gas Price",color = "#9F2C2C",size = 3)+
  geom_text(aes(x = date("2017-06-01"), y = 1.7), label = "Total Riders",color = "#3F4345",size = 3)+
  theme_minimal()+
  theme(axis.title.y.left = element_text(color = "#9F2C2C",size = 12),
        axis.title.y.right = element_text(color = "#3F4345",size = 12),
        axis.text.y.left = element_text(color = "#9F2C2C",size = 8),
        axis.text.y.right = element_text(color = "#3F4345",size = 8),
        axis.title.x.bottom = element_blank(),
        plot.title = element_text(color = "#006bb3",size = 14),
        plot.subtitle = element_text(color = "#006bb3"))+
  facet_wrap(~RouteClass)
```


```{r}
rider_gas_complete %>% 
  filter(RouteClass == "CommExpress") %>% 
  ggplot(aes(x = date, y = NumOfWdays*3,group = Route))+
  geom_line()+
  geom_line(aes(y = riders/1000), color = "blue")
```

